---
title: "Using a custom R package to facilitate teaching"
author: |
  | Mark Andrews
  | Psychology Department, Nottingham Trent University
  | 
  | ```mark.andrews@ntu.ac.uk```
fontsize: 10pt
output:
 beamer_presentation:
  keep_tex: true
  fonttheme: "serif"
  includes:
   in_header: preamble.tex
---

```{r, echo=F}
knitr::opts_chunk$set(echo = T, prompt = F, warning = F, message = F, comment='#>')
# Thanks to 
# https://github.com/ramnathv/slidify/issues/189#issuecomment-15850008
hook1 <- function(x){ gsub("```\n*```r*\n*", "", x) }
hook2 <- function(x){ gsub("```\n+```\n", "", x) }
knitr::knit_hooks$set(document = hook1)
```

```{r, echo=F}
library(tidyverse)
```


# Background: R in NTU Psychology

* For approximately 10 years, we have used R as the software in undergraduate and MSc advanced statistics modules ($\approx 70$ students per year).
* In 2019/20, we began the department wide replacement of SPSS by R as the software used in all statistics and research methods modules.
* In 2019/20, 1st year UG statistics and research methods modules were taught using R ($\approx 1000$ students)
* In 2020/21, 1st and 2nd year UG statistics and research methods modules were taught using R ($\approx 2000$ students)
* In 2021/22, in addition to 1st and 2nd year UG statistics and research methods modules, 3rd year research project data analysis supervision will use R ($\approx 2500$ students)


# Initial motivation for using a custom R package

* By their second workshop in their 1st year, students were being asked to write the likes of the following:
```{r,eval = F, indent = '   '}
Df %>% 
  group_by(identified_as) %>%
  summarise(mean_correct = mean(percent_corr, na.rm = T),
            iqr = IQR(percent_corr, na.rm = T))


ggplot(data = Df, 
       aes(x = Faithful, 
           fill = FaceSex, colour = FaceSex)) +
  geom_histogram(binwidth = .25, 
                 position = "identity", 
                 alpha = .35) +
  labs(x = "Faithfulness rating", 
       y = "Frequency")
```

#  Initial motivation for using a custom R package

* Opaque multiline R code in the first few classes lead many students to conclude that R code is opaque and inscrutable.
* Using R was perceived as a matter of copying and pasting inscrutable code.
* Not much is gained if students replace blindly clicking (in SPSS) by blindly copy-n-pasting (in R).
* Therefore, the objective of a custom R package was primarily to provide custom functions to replace all opaque multiline R code used throughout all workshops and labs.

# Arguments against a custom R package

* Arguments against using a custom R package for teaching were usually variants of the following related arguments:
  
  * students won't learn powerful general tools like `dplyr` and `ggplot`, and instead just learn some limited functions.
  * using custom function in place of general R code means students will not learn R code generally, nor will they understand what calculations the functions implement.
  * students will become dependent on a specific package
  

# Rebuttal to arguments against a custom R package

* The second argument is an argument against all R packages, and even against R (or all high level languages) itself.
* Likewise, the third argument is an argument against all R packages.
* In response to the first argument, while teaching powerful general tools like `dplyr` and `ggplot` are excellent and well worth learning, do we have the resources to properly teach them?

# Overview of `psyntur` (0.0.2)

```{r}
library(psyntur)
packageVersion("psyntur")
```


* The `psyntur`, in its current state, provides
  
  * Some methods for data visualization
  * Some methods for data exploration
  * Some helper functions for statistical analysis
  
* Available at https://github.com/mark-andrews/psyntur
* Package webpage at https://mark-andrews.github.io/psyntur

# Data visualization: Scatterplot

```{r, out.width='0.67\\textwidth', fig.align='center'}
scatterplot(x = trustworthy, y = faithful, 
            data = faithfulfaces)
```


# Data visualization: Coloured scatterplot

```{r, out.width='0.67\\textwidth', fig.align='center'}
scatterplot(x = trustworthy, y = faithful, 
            by = face_sex,
            data = faithfulfaces)
```


# Data visualization: Coloured scatterplot, with line of best fit

```{r, out.width='0.67\\textwidth', fig.align='center'}
scatterplot(x = trustworthy, y = faithful, 
            by = face_sex,
            best_fit_line = TRUE,
            data = faithfulfaces)
```

# Comparing `ggplot` and `psyntur`

* This `scatterplot` function
```{r, eval = F}
scatterplot(x = trustworthy, y = faithful, 
            by = face_sex,
            best_fit_line = TRUE,
            data = faithfulfaces)
```
implements exactly this:
```{r, eval = F}
ggplot(faithfulfaces,
       aes(x = trustworthy, y = faithful, colour = face_sex)
) + geom_point() +
  stat_smooth(method = 'lm', se = FALSE, fullrange = TRUE,
              formula = 'y ~ x') +
  theme_classic() +
  scale_colour_brewer(palette = "Set1")
```


# Exploratory data analysis: Summary statistics

* The `describe` function is a wrapper to `dplyr`'s `summarize` (and `group_by` etc):

```{r}
describe(data = faithfulfaces, 
         avg = mean(faithful), stdev = sd(faithful))
```


```{r}
describe(data = faithfulfaces, 
         by = face_sex, 
         avg = mean(faithful), stdev = sd(faithful))
```

# Exploratory data analysis: Multiple variables, multiple functions

The `describe_across` is an interface to `summarize`, `across`, `pivot_wider`:
```{r}
describe_across(faithfulfaces,
                variables = c(trustworthy, faithful),
                functions = list(avg = mean, 
                                 median = median, 
                                 stdev = sd),
                pivot = TRUE
)
```

# Statistics helper functions
\footnotesize
* We also implemented a few helper functions like the following:
```{r}
shapiro_test(faithful, by = face_sex, 
             data = faithfulfaces)
```
which implements
```{r}
get_shapiro_test_results <- function(x){
    results <- stats::shapiro.test(x)
    tibble(statistic = results$statistic, p_value = results$p.value)
  }
faithfulfaces %>% 
  group_by(face_sex) %>%
  summarise(get_shapiro_test_results(faithful),
            .groups = 'drop')
```


# Other advantages to using a custom R package

* Data files with help files (e.g. `?faithfulfaces`)
* Custom help pages for functions
* Vignettes, see https://mark-andrews.github.io/psyntur/articles/

# The great `install_github` gotcha

* We did have any urgency getting `psyntur` onto CRAN.
* `devtools::install_github` seemed like a perfectly acceptable alternative
* However, hundreds of students hit GitHub's `API rate limit exceeded` restriction in workshops.
* `install_github`, by defaults, asks if all dependencies should be updated to their latest version. This often lead students down the `Do you want to install from sources the package which needs compilation?` rabbit hole. 
